Dynamics of Colombo's Top: Tidal Dissipation and Resonance Capture

### 2021-05-25 writeups/Su_weak_tides.tex
check in new paper draft

    \newcommand*{\scinot}[2]{#1\times10^{#2}}
    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2#1}{\partial#2^2}}
    % inline
    \newcommand*{\mdil}[2]{\mathrm{D}#1/\mathrm{D}#2}
    \newcommand*{\pdil}[2]{\partial#1/\partial#2}
    \newcommand*{\rdil}[2]{\mathrm{d}#1/\mathrm{d}#2}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\left\langle#1\right\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\mathbf{#1}}
    \newcommand*{\uv}[1]{\hat{\mathbf{#1}}}
    \newcommand*{\md}[0]{\mathrm{d}}
    \DeclareMathOperator*{\med}{med}
    \DeclareMathOperator*{\erf}{erf}

@@

In Section~\ref{s:theory}, we briefly review Colombo's Top. In
Section~\ref{s:toy_model}, we investigate the effect of adding a simple
alignment torque to Colombo's Top. The resulting dynamics are illustrative of
the new behavior that emerges due to tidal dissipation. In
Section~\ref{s:full_tide_prob}, we solve for the dynamics of the system
including a realistic model of tidal dissipation. In Section~\ref{s:summary}, we
apply our results to three astrophysical scenarios of interest: (i) a super
Earth with a cold Jupiter companion, (ii) the formation of USPs, and (iii) the
rapid orbital decay of WASP-12b. We summarize and discuss in
Section~\ref{s:summary}.

@@

To understand this process more concretely, as well as associate probabilities
to the two possible outcomes, it is important to be quantitative. The correct
approach is to consider the evolution of the value of the \emph{unperturbed}
Hamiltonian (Eq.~\ref{eq:H}) as the spin evolves due to the alignment torque. A
point in zone III evolves such that $H$ is increasing until $H \approx H_{\rm
sep}$ where $H_{\rm sep}$ is the value of $H$ along the separatrix, given by
\begin{align}
    H_{\rm sep} &\equiv H\p{\cos \theta_{\rm 4}, \phi_{\rm 4}},\nonumber\\
        &\approx g\sin I + \frac{g^2}{2\alpha}\cos^2 I +
            \mathcal{O}\p{\eta^2},\label{eq:def_Hsep}
\end{align}
where $\theta_4$ and $\phi_4$ are the coordinates of CS4. As the system evolves
closer to the separatrix, the change in $H$ over each circulation cycle can be
approximated by $\Delta H_-$, the change in $H$ along $\mathcal{C}_-$ (see
Fig.~\ref{fig:1contours}). In general, we define the quantities $\Delta H_{\pm}$
\begin{equation}
    \Delta H_{\pm} \equiv \oint\limits_{\mathcal{C}_{\pm}}
        \rd{H}{t}\;\mathrm{d}t.\label{eq:def_dHpm}
\end{equation}
This can be simplified by using:
\begin{align}
    \rd{H}{t} &=
            \pd{H}{(\cos \theta)}\rd{(\cos \theta)}{t}
            + \pd{H}{\phi}\rd{\phi}{t},\nonumber\\
        &= \p{\rd{(\cos\theta)}{t}}_{\rm tide} \rd{\phi}{t},\\
    \Delta H_{\pm} &= \mp\frac{1}{t_{\rm al}}
        \int\limits_0^{2\pi} \sin^2\theta\;\mathrm{d}\phi.
\end{align}
In the first line, Hamilton's equations ensure that the non-tidal contributions
to $\rdil{H}{t}$ vanish. Thus, if we evaluate $H$ every time that a trajectory
originating in zone III crosses $\phi = 0$, we see that it will initially be $<
H_{\rm sep}$ and increase for each circulation cycle until the system encounters
the separatrix. At the beginning of the separatrix-crossing orbit, the initial
value of $H$, denoted $H_{\rm i}$, must be greater than $H_{\rm sep} - \Delta H$
to encounter the separatrix on the current orbit, and must be less than $H_{\rm
sep}$, otherwise the trajectory will be in zone II already. We thus have
\begin{equation}
    H_{\rm i} \in \s{ H_{\rm sep} - \Delta H_-,  H_{\rm sep} }
        \label{eq:Hi_range},
\end{equation}
The values of $\cos \theta$ corresponding to the lower and upper bounds on this
range are shown as the black and red dots on the left of Fig.~\ref{fig:6equils}
respectively.

During the separatrix-crossing orbit, the trajectory first evolves
approximately along $\mathcal{C}_-$ and then along $\mathcal{C}_+$, after which
the final value of $H$, denoted $H_{\rm f}$ is approximately equal to
\begin{equation}
    H_{\rm f} = H_{\rm i} + \Delta H_+ + \Delta H_-.
\end{equation}
There can then be two outcomes depending on the value of $H_{\rm f}$:
\begin{itemize}
    \item If $H_{\rm f} > H_{\rm sep}$, then the trajectory enters the
        separatrix, following the red shaded region in Fig.~\ref{fig:6equils},
        and executes a zone III $\to$ II transition.

    \item If $H_{\rm f} < H_{\rm sep}$, then the trajectory exits the separatrix
        above CS4, following the yellow shaded region in Fig.~\ref{fig:6equils}
        and executes a zone III $\to$ I transition.
\end{itemize}
These two possibilites can be re-expressed in terms of $H_{\rm i}$: we find that
if $H_{\rm i}$ is in the interval $\s{H_{\rm sep} - \Delta H_-, H_{\rm sep} -
\Delta H_{-} - \Delta H_+}$, then the system executes a III $\to$ I transition,
and if it is in the interval $\s{H_{\rm sep} - \Delta H_- - \Delta H_+, H_{\rm
sep}}$, then the system executes a III $\to$ II transition. We see that there is
a critical value of $H_{\rm i}$ in the interval given by
Eq.~\eqref{eq:Hi_range}, where $H_{\rm i} = H_{\rm sep} - \Delta H_{-} - \Delta
H_+$, that separates the two possible outcomes of the separatrix encounter
within the interval given by Eq.~\eqref{eq:Hi_range}. The value of $\cos \theta$
for which $H$ is equal to $H_{\rm sep} - \Delta H_{-} - \Delta H_+$ is shown as
the blue dot on the left of Fig.~\ref{fig:6equils}. Finally, if the alignment
torque is weak, then $\Delta H_- \sim \mathcal{O}\p{t_{\rm al}^{-1}}$ is small
compared to any variation in the value of $H$ (e.g.\ when changing $\phi_0$ or
$\theta_0$ by a small amount), and $H_{\rm i}$ can be effectively considered as
randomly chosen from a uniform distribution over $\s{H_{\rm sep} - \Delta H_-,
H_{\rm sep}}$. As a consequence we obtain the probability of a III $\to$ II
transition:
\begin{equation}
    P_{\rm III \to II} = \frac{\Delta H_- + \Delta H_+}{\Delta H_-}.
        \label{eq:def_P32_toy}
\end{equation}

@@

Finally, we remark that the calculation above is just an of \emph{Melnikov's
Method} \citep{g_and_h}. Melnikov's Method is a general calculation that gives
the degree of splitting of a ``homoclinic orbit'' (here, the separatrix) induced
by a small, possibly time-dependent, perturbation. The explicit connection
between the evolution of the unperturbed Hamiltonian and distances between
curves in phase space (as seen in Fig.~\ref{fig:6equils}) is provided by
Melnikov's Method.

@@

Having understood the effect of introducing an additional dissipative torque to
the Colombo's Top system, we now implement a realistic model of tidal
dissipation. We use the weak friction theory of equilibrium tides
\citep{lai2012}. In this model, tides cause both the spin orientation $\uv{s}$
and frequency $\Omega_{\rm s}$ to evolve on the characteristic tidal timescale
$t_{\rm s}$ following:
\begin{align}
    \p{\rd{\uv{s}}{t}}_{\rm tide} &= \frac{1}{t_{\rm s}}
                \s{\frac{2n}{\Omega_{\rm s}} - \p{\uv{s} \cdot \uv{l}}}
                    \uv{s} \times \p{\uv{l} \times \uv{s}}\label{eq:dsdt_tide},\\
    \frac{1}{\Omega_{\rm s}}\p{\rd{\Omega_{\rm s}}{t}}_{\rm tide}
        &= \frac{1}{t_{\rm s}} \s{\frac{2n}{\Omega_{\rm s}}\p{\uv{s} \cdot
            \uv{l}} - 1 - \p{\uv{s} \cdot \uv{l}}^2},\label{eq:dWsdt_tide}
\end{align}
where $t_{\rm s}$ is given by
\begin{equation}
    \frac{1}{t_{\rm s}} \equiv \frac{L}{2S} \frac{\Omega_{\rm
        s}}{2n}\frac{3k_2}{Q}\p{\frac{M_\star}{m}}\p{\frac{R}{a}}^5 n.
\end{equation}
Here, $L = ma^2n$ and $S = kmR^2 \Omega_{\rm s}$ are the orbital and spin
angular momenta of the inner planet, respectively. We neglect orbital evolution
in this section since the time scale is longer than $t_{\rm s}$ by a factor of
$\sim L / S \gg 1$, and so $t_{\rm s}$ is a constant. We will continue mostly
consider the case where tidal dissipation is slow, i.e.\ $\abs{g}t_{\rm s} \gg
1$. The full equations of motion including weak tidal friction can be written in
component form as
\begin{align}
    \rd{\theta}{t} &= g\sin I \sin \phi -
        \frac{1}{t_{\rm s}}\sin \theta\p{\frac{2n}{\Omega_{\rm s}} - \cos \theta}
            ,\label{eq:ds_fullq}\\
    \rd{\phi}{t} &= -\alpha\cos\theta
        - g\p{\cos I + \sin I \cot \theta \cos \phi}\label{eq:ds_fullphi},\\
    \frac{1}{\Omega_{\rm s}}\rd{\Omega_{\rm s}}{t}
        &= \frac{1}{t_{\rm s}} \s{\frac{2n}{\Omega_{\rm s}} \cos \theta
            - \p{1 + \cos^2\theta}}\label{eq:ds_fulls}.
\end{align}
The conditions for which the two tidal terms vanish are given respectively by:
\begin{align}
    \dot{\theta}_{\rm tide} = 0 &: \quad \frac{2n}{\Omega_{\rm s}} = \cos \theta,
        \label{eq:weaktide_dqzero}\\
    \dot{\Omega}_{\rm s} = 0 &: \quad \frac{n}{\Omega_{\rm s}}
        = \frac{1 + \cos^2\theta}{2\cos \theta}.\label{eq:weaktide_dWszero}
\end{align}

@@

Using this result, we can then identify the long-term equilibria of the system
when $\Omega_{\rm s}$ is also evolving: these equilibria of the system must both
satisfy $\dot{\Omega}_{\rm s} = 0$ and be a CS that is stable in the presence of
tidal dissipation (i.e.\ satisfying $\rdil{\uv{s}}{t} = 0$).
The panels of Fig.~\ref{fig:6equils} describes the evolution of
Eqs.~(\ref{eq:dsdt_tide}--\ref{eq:dWsdt_tide}) qualitatively in the coordinates
$(\Omega_{\rm s}, \theta)$, along with the locations of CS1 and CS2, for a few
different system parameters. The circled points in the panels of
Fig.~\ref{fig:6equils} satisfy the criteria to be long-term equilibria, and we
call them \emph{tidal Cassini Equilibria} (tCE). We number tCE1 and tCE2 the tCE
that are in CSs 1 and 2 respectively. The obliquities of the tCE depend on the
system architecture, which can be quantified using the parameter
\begin{align}
    \eta_{\rm sync} &\equiv \big(\eta\big)_{\Omega_{\rm s} = n}
        = \eta \frac{\Omega_{\rm s}}{n}\nonumber\\
        &= \frac{1}{2}\frac{k}{k_{\rm q}}
            \frac{m_{\rm p}m}{M_\star^2}
            \p{\frac{a}{a_{\rm p}}}^3 \p{\frac{a}{R}}^3\cos(I).
            \label{eq:def_etasync}
\end{align}
The tCE obliquity as a function of $\eta_{\rm sync}$ are shown for $I =
20^\circ$ in the top panel of Fig.~\ref{fig:probs20} and for $I = 5^\circ$ in
the top panel of Fig.~\ref{fig:probs5}. In fact, an analytical expression for
the tCE2 obliquity for $\eta_{\rm sync} \ll \eta_{\rm c}$ can be computed using
Eqs.~(\ref{eq:weaktide_dWszero}--\ref{eq:def_etasync}) and that $\cos \theta_2
\approx \eta \cos I$ to yield:
\begin{align}
    \cos \theta_{\rm tCE2} &\approx \sqrt{\frac{\eta_{\rm sync} \cos I}{2}},
        &
    \frac{\Omega_{\rm s, tCE2}}{n} &\approx
        \sqrt{2\cos I \eta_{\rm sync}}. \label{eq:def_tce2_approx}
\end{align}
This approximation for $\theta_{\rm tCE2}$ is shown as the black dashed line in
the top panels of Fig.~\ref{fig:probs20} and~\ref{fig:probs5}, where good
agreement is observed.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/6equils.png}
    \caption{Schematic depiction of the effect of tidal friction on the planet's
    spin for $I = 20^\circ$ and three values of $\eta_{\rm sync}$ (labeled). The
    black and blue lines denote where the tidal $\dot{\Omega}_{\rm s}$ and
    $\dot{\theta}$ change signs
    (Eqs.~\ref{eq:weaktide_dqzero}--\ref{eq:weaktide_dWszero}). The orange and
    green lines give the CS1 and CS2 obliquities respectively, which are the two
    CSs that are stable under the effect of tidal dissipation. Note that when
    $\dot{\theta}_{\rm tide} > 0$, CS1 becomes unstable, denoted by the dashed
    orange line. The points that are both CSs and satisfy $\dot{\Omega}_{\rm s}
    = 0$ are the tidal Cassini Equilibria (tCE), which are circled and labeled.
    }\label{fig:6equils}
\end{figure}

There are two important conditions that can change the existence and stability
of the tCE\@. First, if $\eta_{\rm sync} > \eta_{\rm c}$, then tCE1 will not
exist, as can be seen in the bottom panel of
Fig.~\ref{fig:6equils}\footnote{Strictly speaking, $\eta_{\rm sync}$ can be
slightly smaller than $\eta_{\rm c}$, as the planet's spin is slightly
subsynchronous at tCE1 when $\eta \approx \eta_{\rm c}$, see the middle panel of
Fig.~\ref{fig:6equils}.}. Secondly, tCE2 may not be stable if the tidal phase
shift is too large. Applying the results of Section~\ref{ss:tidal_equils}, we
find that tCE2 is stable as long as $t_{\rm s} \geq t_{\rm s, c}$ where
\begin{equation}
    t_{\rm s, c} \equiv \frac{\sin \theta_{\rm tCE2}}{\abs{g} \sin
            I}\p{\frac{2n}{\Omega_{\rm s, tCE2}} - \cos \theta_{\rm tCE2}}.
\end{equation}
The tCE2 subscripts denote evaluation at tCE2. When $\eta_{\rm sync} \ll
\eta_{\rm c}$, we can use Eq.~\eqref{eq:def_tce2_approx} to further simplify:
\begin{equation}
    t_{\rm s, c}\approx \frac{1}{\abs{g}\sin I}\sqrt{\frac{2}{\eta_{\rm sync}
        \cos I}}. \label{eq:def_ts_crit}
\end{equation}

>>>>

    \newcommand*{\scinot}[2]{#1\times10^{#2}}
    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2#1}{\partial#2^2}}
    % inline
    \newcommand*{\mdil}[2]{\mathrm{D}#1/\mathrm{D}#2}
    \newcommand*{\pdil}[2]{\partial#1/\partial#2}
    \newcommand*{\rdil}[2]{\mathrm{d}#1/\mathrm{d}#2}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\left\langle#1\right\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\mathbf{#1}}
    \newcommand*{\uv}[1]{\hat{\boldsymbol{\mathbf{#1}}}}
    \newcommand*{\md}[0]{\mathrm{d}}
    \DeclareMathOperator*{\med}{med}
    \DeclareMathOperator*{\erf}{erf}

@@

Our paper is organized as follows. In Section~\ref{s:theory}, we briefly review
Colombo's Top. In Section~\ref{s:toy_model}, we investigate the effect of adding
a simple alignment torque to Colombo's Top. The resulting dynamics are
illustrative of the new behavior that emerges due to tidal dissipation. In
Section~\ref{s:full_tide_prob}, we solve for the dynamics of the system
including a realistic model of tidal dissipation. In Section~\ref{s:summary}, we
apply our results to three astrophysical scenarios of interest: (i) a super
Earth with a cold Jupiter companion, (ii) the formation of USPs, and (iii) the
rapid orbital decay of WASP-12b. We summarize and discuss in
Section~\ref{s:summary}.

@@

To understand this process more concretely, as well as associate probabilities
to the two possible outcomes, it is important to be quantitative. The correct
approach is to consider the evolution of the value of the \emph{unperturbed}
Hamiltonian (Eq.~\ref{eq:H}) as the spin evolves due to the alignment torque. A
point in zone III evolves such that $H$ is increasing until $H \approx H_{\rm
sep}$ where $H_{\rm sep}$ is the value of $H$ along the separatrix, given by
\begin{align}
    H_{\rm sep} &\equiv H\p{\cos \theta_{\rm 4}, \phi_{\rm 4}},\nonumber\\
        &\approx g\sin I + \frac{g^2}{2\alpha}\cos^2 I +
            \mathcal{O}\p{\eta^2},\label{eq:def_Hsep}
\end{align}
where $\theta_4$ and $\phi_4$ are the coordinates of CS4. As the system evolves
closer to the separatrix, the change in $H$ over each circulation cycle can be
approximated by $\Delta H_-$, the change in $H$ along $\mathcal{C}_-$ (see
Fig.~\ref{fig:1contours}). In general, we define the quantities $\Delta H_{\pm}$
\begin{equation}
    \Delta H_{\pm} \equiv \oint\limits_{\mathcal{C}_{\pm}}
        \rd{H}{t}\;\mathrm{d}t.\label{eq:def_dHpm}
\end{equation}
This can be simplified by using:
\begin{align}
    \rd{H}{t} &=
            \pd{H}{(\cos \theta)}\rd{(\cos \theta)}{t}
            + \pd{H}{\phi}\rd{\phi}{t},\nonumber\\
        &= \p{\rd{(\cos\theta)}{t}}_{\rm tide} \rd{\phi}{t},\\
    \Delta H_{\pm} &= \mp\frac{1}{t_{\rm al}}
        \int\limits_0^{2\pi} \sin^2\theta\;\mathrm{d}\phi.
\end{align}
In the first line, Hamilton's equations ensure that the non-tidal contributions
to $\rdil{H}{t}$ vanish. Thus, if we evaluate $H$ every time that a trajectory
originating in zone III crosses $\phi = 0$, we see that it will initially be $<
H_{\rm sep}$ and increase for each circulation cycle until the system encounters
the separatrix. At the beginning of the separatrix-crossing orbit, the initial
value of $H$, denoted $H_{\rm i}$, must be greater than $H_{\rm sep} - \Delta H$
to encounter the separatrix on the current orbit, and must be less than $H_{\rm
sep}$, otherwise the trajectory will be in zone II already. We thus have
\begin{equation}
    H_{\rm i} \in \s{ H_{\rm sep} - \Delta H_-,  H_{\rm sep} }
        \label{eq:Hi_range},
\end{equation}
The values of $\cos \theta$ corresponding to the lower and upper bounds on this
range are shown as the black and red dots on the left of
Fig.~\ref{fig:toy_hop_manifolds} respectively.

During the separatrix-crossing orbit, the trajectory first evolves
approximately along $\mathcal{C}_-$ and then along $\mathcal{C}_+$, after which
the final value of $H$, denoted $H_{\rm f}$ is approximately equal to
\begin{equation}
    H_{\rm f} = H_{\rm i} + \Delta H_+ + \Delta H_-.
\end{equation}
There can then be two outcomes depending on the value of $H_{\rm f}$:
\begin{itemize}
    \item If $H_{\rm f} > H_{\rm sep}$, then the trajectory enters the
        separatrix, following the red shaded region in
        Fig.~\ref{fig:toy_hop_manifolds}, and executes a zone III $\to$ II
        transition.

    \item If $H_{\rm f} < H_{\rm sep}$, then the trajectory exits the separatrix
        above CS4, following the yellow shaded region in
        Fig.~\ref{fig:toy_hop_manifolds} and executes a zone III $\to$ I
        transition.
\end{itemize}
These two possibilites can be re-expressed in terms of $H_{\rm i}$: we find that
if $H_{\rm i}$ is in the interval $\s{H_{\rm sep} - \Delta H_-, H_{\rm sep} -
\Delta H_{-} - \Delta H_+}$, then the system executes a III $\to$ I transition,
and if it is in the interval $\s{H_{\rm sep} - \Delta H_- - \Delta H_+, H_{\rm
sep}}$, then the system executes a III $\to$ II transition. We see that there is
a critical value of $H_{\rm i}$ in the interval given by
Eq.~\eqref{eq:Hi_range}, where $H_{\rm i} = H_{\rm sep} - \Delta H_{-} - \Delta
H_+$, that separates the two possible outcomes of the separatrix encounter
within the interval given by Eq.~\eqref{eq:Hi_range}. The value of $\cos \theta$
for which $H$ is equal to $H_{\rm sep} - \Delta H_{-} - \Delta H_+$ is shown as
the blue dot on the left of Fig.~\ref{fig:toy_hop_manifolds}. Finally, if the
alignment torque is weak, then $\Delta H_- \sim \mathcal{O}\p{t_{\rm al}^{-1}}$
is small compared to any variation in the value of $H$ (e.g.\ when changing
$\phi_0$ or $\theta_0$ by a small amount), and $H_{\rm i}$ can be effectively
considered as randomly chosen from a uniform distribution over $\s{H_{\rm sep} -
\Delta H_-, H_{\rm sep}}$. As a consequence we obtain the probability of a III
$\to$ II transition:
\begin{equation}
    P_{\rm III \to II} = \frac{\Delta H_- + \Delta H_+}{\Delta H_-}.
        \label{eq:def_P32_toy}
\end{equation}

@@

Finally, we remark that the calculation above is just an of \emph{Melnikov's
Method} \citep{g_and_h}. Melnikov's Method is a general calculation that gives
the degree of splitting of a ``homoclinic orbit'' (here, the separatrix) induced
by a small, possibly time-dependent, perturbation. The explicit connection
between the evolution of the unperturbed Hamiltonian and distances between
curves in phase space (as seen in Fig.~\ref{fig:toy_hop_manifolds}) is provided by
Melnikov's Method.

@@

Having understood the effect of introducing an additional dissipative torque to
the Colombo's Top system, we now implement a realistic model of tidal
dissipation. We use the weak friction theory of equilibrium tides
\citep{lai2012}. In this model, tides cause both the spin orientation $\uv{s}$
and frequency $\Omega_{\rm s}$ to evolve on the characteristic tidal timescale
$t_{\rm s}$ following:
\begin{align}
    \p{\rd{\uv{s}}{t}}_{\rm tide} &= \frac{1}{t_{\rm s}}
                \s{\frac{2n}{\Omega_{\rm s}} - \p{\uv{s} \cdot \uv{l}}}
                    \uv{s} \times \p{\uv{l} \times \uv{s}}\label{eq:dsdt_tide},\\
    \frac{1}{\Omega_{\rm s}}\p{\rd{\Omega_{\rm s}}{t}}_{\rm tide}
        &= \frac{1}{t_{\rm s}} \s{\frac{2n}{\Omega_{\rm s}}\p{\uv{s} \cdot
            \uv{l}} - 1 - \p{\uv{s} \cdot \uv{l}}^2},\label{eq:dWsdt_tide}
\end{align}
where $t_{\rm s}$ is given by
\begin{equation}
    \frac{1}{t_{\rm s}} \equiv \frac{1}{4k}
        \frac{3k_2}{Q}\p{\frac{M_\star}{m}}\p{\frac{R}{a}}^3 n.
\end{equation}
In this section, we neglect orbital evolution in this section since the time
scale is longer than $t_{\rm s}$ by a factor of $\sim L / S \gg 1$, and so
$t_{\rm s}$ is a constant. We will continue mostly consider the case where tidal
dissipation is slow, i.e.\ $\abs{g}t_{\rm s} \gg 1$. The full equations of
motion including weak tidal friction can be written in component form as
\begin{align}
    \rd{\theta}{t} &= g\sin I \sin \phi -
        \frac{1}{t_{\rm s}}\sin \theta\p{\frac{2n}{\Omega_{\rm s}} - \cos \theta}
            ,\label{eq:ds_fullq}\\
    \rd{\phi}{t} &= -\alpha\cos\theta
        - g\p{\cos I + \sin I \cot \theta \cos \phi}\label{eq:ds_fullphi},\\
    \frac{1}{\Omega_{\rm s}}\rd{\Omega_{\rm s}}{t}
        &= \frac{1}{t_{\rm s}} \s{\frac{2n}{\Omega_{\rm s}} \cos \theta
            - \p{1 + \cos^2\theta}}\label{eq:ds_fulls}.
\end{align}
The conditions for which the two tidal terms vanish are given respectively by:
\begin{align}
    \dot{\theta}_{\rm tide} = 0 &: \quad \frac{2n}{\Omega_{\rm s}} = \cos \theta,
        \label{eq:weaktide_dqzero}\\
    \dot{\Omega}_{\rm s} = 0 &: \quad \frac{n}{\Omega_{\rm s}}
        = \frac{1 + \cos^2\theta}{2\cos \theta}.\label{eq:weaktide_dWszero}
\end{align}

@@

Using this result, we can then identify the long-term equilibria of the system
when $\Omega_{\rm s}$ is also evolving: these equilibria of the system must both
satisfy $\dot{\Omega}_{\rm s} = 0$ and be a CS that is stable in the presence of
tidal dissipation (i.e.\ satisfying $\rdil{\uv{s}}{t} = 0$).
Figure~\ref{fig:6equilssingle} describes the evolution of
Eqs.~(\ref{eq:dsdt_tide}--\ref{eq:dWsdt_tide}) qualitatively in the coordinates
$(\Omega_{\rm s}, \theta)$, along with the locations of CS1 and CS2. The circled
points in the panels of Fig.~\ref{fig:6equils} satisfy the criteria to be
long-term equilibria, and we call them \emph{tidal Cassini Equilibria} (tCE). We
number tCE1 and tCE2 the tCE that are in CSs 1 and 2 respectively. The
obliquities of the tCE depend on the system architecture, which can be
quantified using the parameter
\begin{align}
    \eta_{\rm sync} &\equiv \big(\eta\big)_{\Omega_{\rm s} = n}
        = \eta \frac{\Omega_{\rm s}}{n}\nonumber\\
        &= \frac{1}{2}\frac{k}{k_{\rm q}}
            \frac{m_{\rm p}m}{M_\star^2}
            \p{\frac{a}{a_{\rm p}}}^3 \p{\frac{a}{R}}^3\cos(I).
            \label{eq:def_etasync}
\end{align}
In Fig.~\ref{fig:6equilssingle}, $\eta_{\rm sync} = 0.06$; the two panels of
Fig.~\ref{fig:6equils} illustrate the change to the tCE when $\eta_{\rm sync} =
0.5$ and $0.7$ respectively.

The tCE obliquity as a function of $\eta_{\rm sync}$ are shown for $I =
20^\circ$ in the top panel of Fig.~\ref{fig:probs20} and for $I = 5^\circ$ in
the top panel of Fig.~\ref{fig:probs5}. In fact, an analytical expression for
the tCE2 obliquity for $\eta_{\rm sync} \ll \eta_{\rm c}$ can be computed using
Eqs.~(\ref{eq:weaktide_dWszero}--\ref{eq:def_etasync}) and that $\cos \theta_2
\approx \eta \cos I$ to yield:
\begin{align}
    \cos \theta_{\rm tCE2} &\approx \sqrt{\frac{\eta_{\rm sync} \cos I}{2}},
        &
    \frac{\Omega_{\rm s, tCE2}}{n} &\approx
        \sqrt{2\cos I \eta_{\rm sync}}. \label{eq:def_tce2_approx}
\end{align}
This approximation for $\theta_{\rm tCE2}$ is shown as the black dashed line in
the top panels of Fig.~\ref{fig:probs20} and~\ref{fig:probs5}, where good
agreement is observed.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{../initial/1_weaktide/6equilssingle.png}
    \caption{Schematic depiction of the effect of tidal friction on the planet's
    spin for $I = 20^\circ$, corresponding to $\eta_{\rm c} \approx 0.574$
    (Eq.~\ref{eq:def_etac}), and $\eta_{\rm sync} = 0.06$. The black and blue
    lines denote where the tidal $\dot{\Omega}_{\rm s}$ and $\dot{\theta}$
    change signs (Eqs.~\ref{eq:weaktide_dqzero}--\ref{eq:weaktide_dWszero}). The
    orange and green lines give the CS1 and CS2 obliquities respectively, which
    are the two CSs that are stable under the effect of tidal dissipation. Note
    that when $\dot{\theta}_{\rm tide} > 0$, CS1 becomes unstable, denoted by
    the dashed orange line. The points that are both CSs and satisfy
    $\dot{\Omega}_{\rm s} = 0$ are the tidal Cassini Equilibria (tCE), which are
    circled and labeled. The various colored crosses and their associated
    colored lines represent a few characteristic examples of the evolution of
    Colombo's Top under weak tidal friction (for illustrative purposes, we have
    used $\abs{g}t_{\rm s} = 10^2$). The phase space evolution of the two
    thicker evolutionary trajectories (cyan and pink; those beginning at
    $\theta_{\rm i} = 120^\circ$) are shown in
    Figs.~\ref{fig:trajs1}--\ref{fig:trajs2}.
    }\label{fig:6equilssingle}
\end{figure*}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/6equils.png}
    \caption{Same as Fig.~\ref{fig:6equilssingle} but for $\eta_{\rm sync} =
    0.5$ and $\eta_{\rm sync} = 0.7$ in the two panels (labeled). Note in the
    bottom panel that $\eta_{\rm sync} = 0.7 > \eta_{\rm c}$ and tCE1 does not
    exist. In both panels, the light-colored crosses and lines correspond to
    evolutionary trajectories using the same initial conditions as those shown
    in Fig.~\ref{fig:6equilssingle}. The phase space evolution of the thick
    purple trajectory in the bottom panel (starting at $\theta_{\rm i} \approx
    10^\circ$) is shown in Fig.~\ref{fig:trajs3}. }\label{fig:6equils}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/6equils_sims/0_06_2.png}
    \caption{Phase space evolution of the trajectory in
    Fig.~\ref{fig:6equilssingle} colored in pink, for which $\eta_{\rm sync} =
    0.06$ and $I = 20^\circ$. In the
    left-most panel, the trajectory's evolution in $\p{\Omega_{\rm s}, \theta}$
    coordinates along with the curves indicating CS1, 2, and $\dot{\Omega}_{\rm
    s} = 0$ are re-displayed from Fig.~\ref{fig:6equilssingle}. The vertical
    dashed lines denote the values of $\Omega_{\rm s}$ for which the phase space
    evolution of the system is displayed in the right four panels. In each of
    these right four panels, the trajectory's evolution for a single
    circulation/libration cycle is displayed in $\p{\cos \theta, \phi}$ phase
    space for the labeled value of $\Omega_{\rm s}$. The system encounters the
    separatrix, undergoes a III $\to$ I transition, and converges to tCE1.
    }\label{fig:trajs1}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/6equils_sims/0_06_3.png}
    \caption{Same as Fig.~\ref{fig:trajs1} but for the trajectory in
    Fig.~\ref{fig:6equilssingle} colored in cyan, for which $\eta_{\rm sync} =
    0.06$ and $I = 20^\circ$. The system encounters the
    separatrix, undergoes a III $\to$ II transition, and converges to tCE2. Note
    that the initial condition of this system and that displayed in
    Fig.~\ref{fig:trajs1} have the same initial $\theta_{\rm i}$ and
    $\Omega_{\rm s, i}$ and only different precessional phases $\phi_{\rm i}$.
    }\label{fig:trajs2}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/6equils_sims/0_70_0.png}
    \caption{Same as Fig.~\ref{fig:trajs1} but for the purple trajectory
    shown in the bottom panel of Fig.~\ref{fig:6equils}, for which $\eta_{\rm
    sync} = 0.7$ and $I = 20^\circ$. Here, the system
    evolves along CS1 until the separatrix disappears, upon which it experiences
    large obliquity variations that damp due to tidal dissipation. The system
    finally converges to tCE2, the only tCE that exists for this value of
    $\eta_{\rm sync}$ (see bottom panel of Fig.~\ref{fig:6equils}).}\label{fig:trajs3}
\end{figure}

There are two important conditions that can change the existence and stability
of the tCE\@. First, if $\eta_{\rm sync} > \eta_{\rm c}$, then tCE1 will not
exist, as can be seen in the bottom panel of
Fig.~\ref{fig:6equils}\footnote{Strictly speaking, $\eta_{\rm sync}$ can be
slightly smaller than $\eta_{\rm c}$, as the planet's spin is slightly
subsynchronous at tCE1 when $\eta \approx \eta_{\rm c}$, see the top panel of
Fig.~\ref{fig:6equils}.}. Secondly, tCE2 may not be stable if the tidal phase
shift is too large. Applying the results of Section~\ref{ss:tidal_equils}, we
find that tCE2 is stable as long as $t_{\rm s} \geq t_{\rm s, c}$ where
\begin{equation}
    t_{\rm s, c} \equiv \frac{\sin \theta_{\rm tCE2}}{\abs{g} \sin
            I}\p{\frac{2n}{\Omega_{\rm s, tCE2}} - \cos \theta_{\rm tCE2}}.
\end{equation}
The tCE2 subscripts denote evaluation at tCE2. When $\eta_{\rm sync} \ll
\eta_{\rm c}$, we can use Eq.~\eqref{eq:def_tce2_approx} to further simplify:
\begin{equation}
    t_{\rm s, c}\approx \frac{1}{\abs{g}\sin I}\sqrt{\frac{2}{\eta_{\rm sync}
        \cos I}}. \label{eq:def_ts_crit}
\end{equation}
Note that this result differs in form from that given in
\citet{fabrycky_otides}, which assume that the orbital precession of $\uv{l}$ is
driven by the central star. Our expression is simpler and is applicable to
systems where the orbital precession of the inner planet is dominated by an
exterior planet.
